---
title: "Cancer Study"
author: "Dillon Hector"
date: "3/25/2020"
output: html_notebook
---

Does cancer and death rate due to cancer change between US States. I am going to research this simple question to learn if there is a difference and if so, does population account for the change? This topic is important personally because many of my family memebers have died due to cancer and I am interested in learning more about the topic.

```{r}
#Adding the needed packages

library(DataComputing)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(party)

```

```{r}
#Importing the needed datasets

cancerincedents <- read.csv(file.choose(), header = T)
cancerdeaths <- read.csv(file.choose(), header = T)
statepop <- read.csv(file.choose(), header = T)
stateabbrev <- read.csv(file.choose(), header = T)

```
```{r}
#Inspecting the data that was imported

glimpse(cancerincedents)
glimpse(cancerdeaths)
glimpse(statepop)
glimpse(stateabbrev)

head(cancerincedents)
head(cancerdeaths)
head(statepop)
head(stateabbrev)
```

```{r}
#The first task that needs to be done is to wranglet the datasets togetheer into one dataset

#The first wrangle will be to join stateabbrev with statepop

statepopcorrected <-
  statepop %>%
  left_join(stateabbrev, by = "State") 

#Now I will delete the variable "Abbrev" 

statepopcorrected2 <- statepopcorrected[,-3]

#Here is the resulting table

statepopcorrected2
```


```{r}
#I am not fond with the variable name "X2018.Population" so I will change that now

FinalStatepop <-
       statepopcorrected2 %>%
       rename(StatePop = X2018.Population)

FinalStatepop
  
```

```{r}
#Now I am going to rename the variables in cancer incidents and deaths to be more informative

CancerIncedents2 <-
  cancerincedents %>%
  rename(cancerrate = rate) %>%
  select(state, cancerrate)

CancerIncedents2

```
```{r}
#I am now going to do the same task I did above but with cancerdeaths dataset

CancerDeaths2 <-
  cancerdeaths %>%
  rename(deathrate = rate) %>%
  select(state, deathrate)

CancerDeaths2

```
```{r}
#I am now going to join the CancerIncidents2 and CancerDeaths2 dataset together

CancerData <-
  CancerIncedents2 %>%
  left_join(CancerDeaths2, by = "state")

CancerData

```
```{r}
#Finally we will join the CancerData with the FinalStatePop Data

Cancer <-
  FinalStatepop %>%
  select(State, StatePop, Code) %>%
  left_join(CancerData, by = c("Code" = "state"))

Cancer

```

```{r}
#I am now going to create a new variable that gives the percentage of deathrate

Cancer <-
Cancer %>%
  mutate(deathpercent = deathrate/cancerrate) %>%
  arrange(desc(deathpercent))
Cancer

```

```{r}
Cancer %>%
  summarise(MeanCancer = mean(cancerrate), MeanDeaths = mean(deathrate)) %>%
  mutate(meandeathrate = MeanDeaths/MeanCancer)

```

We can see above that Nevada has the highest deaths per cancer patient. This is odd seeing how the population is not the largest by far and the cancer rate is below the mean cancer rate across all states which is shown above. The death rate is about average though which is probably why the percentage is so high.  

```{r}
#I am going to round the numbers in death percent 

Cancer <-
  Cancer %>%
  mutate(deathpercent = round(deathpercent, digits = 2))
  
Cancer

```

```{r}
#I am now going to create a dataframe that will let me make me see the states with each death percent

Cancer2 <-
  Cancer %>%
  select(State, deathpercent) %>%
  arrange(desc(deathpercent))

Cancer2


```

```{r}
#Now that we have the new data table I am going to spread the table with State

Cancer2 <- 
  Cancer2 %>%
  spread(key = State, value = deathpercent) 
Cancer2
```

Looking above we can easily see the death percents for each state. The table has been layed out so you can see the information very easily for your chosen state. The states are also in alphabetical order now so it is also easier to find your state.


```{r}
#Now we will look at some plots to see if we can see a relationship between the variables
#First we will look at State vs Cancer rate plot

Cancer %>%
  ggplot(aes(x = cancerrate, y = deathrate)) + 
  geom_point()
  

  

```

We can see above that as the cancer rate increases so does the death rate. There is a noticeable linear relationship between the two.

```{r}
#Now I am going to add population into the graph to see if population effects the cancer and/or death rate.

Cancer %>%
  ggplot(aes(x = cancerrate, y = deathrate)) + 
  geom_point(aes(size = StatePop))

```

Although there are a few outliers we can see that most states fall into the middle range of population and that most of the larger populated states fall into the lower end of cancer and death rate.

```{r}
#Now I am going to add State into the graph to specify which states are the outliers

Cancer %>%
  ggplot(aes(x = cancerrate, y = deathrate)) + 
  geom_point(aes(size = StatePop, alpha = .2)) +
  geom_text(aes(label = Code)) +
  xlab("Cancer Rate per 100,000 people") +
  ylab("Death rates from Cancer")

```

We can see on the graph above that the outlier that was talked about in the previous segment is in fact Kentucky which has a small population but a high cancer rate and death rate. We can see that most of the states fall between the 440-480 Cancer rate range and that most states also fall between 150-180 cancer related deaths. There are some outliers but I don't really think there is enough evidence to assume that different states have a dramtically different cancer rate and the population count also doesn't seem to effect the rate either.


To understand the relationship better we will make a ctree with death rate as the ending variable.


```{r}
ctreefunc <- party::ctree(deathrate ~ cancerrate + StatePop, data = Cancer)

plot(ctreefunc, type = "simple")
```

We can clearly see above that as the cancer rate raises the deathrate due to cancer is greater, which is understandable. Although looking above to the Death Rates from cancer and cancer rate table we can see there are some outliers such as conneticut which has a cancer rate above 450 but has a low death rate compared to pther states in that cancer rate area with only about 140 deaths.

